plyranges: a grammar for manipulating genomics data

### 2018-02-11 paper.Rmd
initial draft of design and implemenation, needs more examples

Discussion of the API and its key features

- every action on a Range is a verb (functional composition, method chaining)
- fluent (human readable, code describes what to do rather than 'how')
- every function returns a class familiar to the user
- an expressive algebra for performing arithmetic (anchoring functions)
- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)
- recasting overlapping/nearest etc as joins

>>>>

The _plyranges_ API implements a domain specific language using the existing
_IRanges_ and _GenomicRanges_ packages in Bioconductor as a backend. Consequently,
our API still has the speed and effeciency of the aforementioned packages but
with a more coherent interface. The API also
extends the grammar elements in _dplyr_ for performing genomic specific
manipulations such as finding overlapping regions or nearest neighbour regions
between many _Ranges_. The _plyranges_ API is specifically designed to enable
fast interactive analysis of _Ranges_ objects but can also be used for scripting 
genomic data workflows.

We have desgined the API to be fluent. Every function call corresponds to an
action on a _Ranges_ object (they are named verbs) and where possible
functions have few arguments. Each verb is constructed
to enable a tab completion based workflow. Both of these aspects reduce the 
cognitive load on a new user  since most manipulations can be performed with a 
vocabulary of several verbs, rather than having to memorise functions with many 
arguments that are nouns (as is required in the existing Bioconductor packages). 
This is also has the advantage of allowing users to write human readable code
because verbs describe what the code is doing rather than how its doing it.

Workflows can be composed by chaining verbs together via the pipe operator,`%>%` 
(exported from the R package _magrittr_). This is possible because every 
function call is endomorphic: when the input is _Ranges_ object the output
will also be a _Ranges_ object. One advantage of this static typing is that it 
does not  require any additional learning of classes beyond _Ranges_ and 
the _DataFrame_ classes. This is similar to the bedtools API, where the
output is usually a BED file. However, it strongly deviates from the design of 
the _Ranges_ Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables effecient computing as users
are exposed to low-level features of its API which _plyranges_ trys to abstract
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

## Working with Ranges

The _plyranges_ API exports the six core verbs from the _dplyr_ package
and modifys them for use with simple _Ranges_ and _Genomic Ranges_. The
verb `mutate()` takes a Ranges object and a set of name-value pairs
and generates a new Ranges object that with modified or new metadata
columns or modified core components (start, end, width, seqnames, strand). 
The `filter()` function takes a Ranges object and logical expressions and
restricts Ranges object to where the logical expression evaluates to true.
The `summarise()` function takes a Ranges object and a set of name-value pairs
and aggregates the Ranges according to functions evaluated in the name-value pairs.
As `summarise()` is an aggregation it may break the structure of the of a Ranges
object, hence it returns a DataFrame object. The `select()` function determines
which metadata columns are returned and the order they are returned in. The 
`arrange()` function sorts a Ranges object by named variables.
The `group_by()` function creates an implicit grouping of Ranges object according
to variables in the Ranges object. This modifies the actions of `mutate()`,
`summarise()` and `filter()` (and also the set operations, `reduce_ranges()`
and `disjoin_ranges()`), so they are performed on each partition created by the
grouping.

The API has an expressive algebra for performing arithmetic on Ranges via
the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
modifies the width of a Ranges object, while `stretch` extends the start and
end of a Ranges object. These can be chained with the anchoring functions
`anchor_start()`, `anchor_end()`, `anchor_center()`, `anchor_3p()` or 
`anchor_5p()`, which fix the coordinates of a Ranges object in place. 
Moreover, the `shift_` and `flank_` family of functions can be used to shift
all coordinates in a Ranges object or generate flanking regions from a Ranges 
object to the left, right, upstream or downstream of the input. Unlike, the
Bioconductor API, _plyranges_ makes it expclit via function calls whether to
take into account the strand information of a _Ranges_ object.

- recasting overlapping/nearest etc as joins

- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)

- note on the NSE

# Results

# Availablilty and Future Work

The _plyranges_ package is avaiable on the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or can be accessed
via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
We aim to continue developing the `plyranges` package and extend it for use
with more complex data structures outside of genomics such as the 
_SummarizedExperiment_ class. As the `plyranges` interface encourages tidy data 
practices it also integrates well with the grammar of graphics, we aim to use
it for the visualisation of genomic ranges.
